---
title: "Moderation and Mediation"
subtitle: "PSY9003"
date: "2020/03/31"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "css/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: 16:9
      beforeInit: "js/macros.js"
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
#library(DiagrammeR)
library(diagram)
library(psych)
library(psychTools)
library(ggdag)
library(medmod)
library(interactions)
library(processR)
data(epi.bfi)
#data(Tal_Or)
estress <- read_csv("data/estress.csv")
```

# Mediation and moderation

In linear regression, we're looking to understand the relationship between 
*predictors* and *outcomes*.
.pull-left[
```{r examp-lm, echo = FALSE, fig.height = 5, message = FALSE, fig.width= 6, fig.show = "hide"}
set.seed(200)
theme_set(theme_bw())
a <- rnorm(300)
X <- rnorm(300)
Y <- a + X * 0.5 + 0.5 * rnorm(300)
qplot(X, Y) +
  theme(text = element_text(size = 18)) + 
  stat_smooth(method = "lm")
X1 <- rnorm(300)
X2 <- rnorm(300)
```
![](`r knitr::fig_chunk("examp-lm", "png")`)
]
.pull-right[
```{r echo = FALSE}
openplotmat(asp = 1)
textrect(c(.1, .5), .1, lab = "X")
textrect(c(.9, .5), .1, lab = "Y")
straightarrow(c(.2, .5), c(.8, .5))
```
]

---
class: inverse, center, middle
# Moderation

---
# Moderation 

Moderation changes the strength of the relationship between the predictor and dependent variable. 
.center[
```{r echo = FALSE}
pmacroModel(1)
```
]

---
# Moderation
.center[
```{r echo = FALSE}
statisticalDiagram(1)
```
]

---
# The *epi.bfi* dataset

The `epi.bfi` dataset from the `psychTools` package

```{r}
head(epi.bfi)
```

---
# Simple linear regression

Suppose that we want to model `bdi` (Beck Depression Inventory) as a function of `stateanx` (State Anxiety)

```{r}
st_bdi <- lm(bdi ~ stateanx, data = epi.bfi)
summary(st_bdi)
```

---
# Multiple linear regression

An additional predictor that we may find interesting is `epiNeur` - a measure of *neuroticism* from the *Eysenck Personality Inventory*. We can add this to make a *multiple* regression model.

```{r}
st_neu <- lm(bdi ~ stateanx + epiNeur, data = epi.bfi)
summary(st_neu)
```

---
# Adding interaction terms

What if the effect of `stateanx` depends on the level of `epiNeur`? For example, people who score high on *neuroticism* might be more affected by *state anxiety* than people who are low on *neuroticism*. 

We add both main effects and interactions by using `*` instead of `+`

Thus, `stateanx * epiNeur` will give us the main effect of `stateanx`, the main effect of `epiNeur`, and the interaction between the two.

```{r}
int_model <- lm(bdi ~ stateanx * epiNeur, data = epi.bfi)
```

---

```{r highlight.output=14}
summary(int_model)
```

---
# Interpreting the coefficients

```{r}
coef(int_model)
```

The coefficients tell you what the effect of a 1 unit increase in the variable has on the dependent variable.

The coefficients of the main effects (`stateanx` and `epiNeur`) are hard to interpret *in the presence of an interaction* unless the variables have been **mean-centred**. 

---
# Simple slopes

.pull-left[
The `interact_plot()` function from the `interactions` package provides a nice way to visualize this interaction.

```{r simp-slope, fig.height=5, fig.show = "hide"}
interact_plot(int_model,
              pred = stateanx,
              modx = epiNeur)
```
]
.pull-right[
![](`r knitr::fig_chunk("simp-slope", "png")`)
]

---
# Simple slopes

.pull-left[
We can also add individual data points using `plot.points = TRUE`.

Confidence intervals can be added using `interval = TRUE`.

```{r simp-points, fig.height=5, fig.show = "hide"}
interact_plot(int_model,
              pred = stateanx,
              modx = epiNeur,
              plot.points = TRUE,
              interval = TRUE)
```
]
.pull-right[
![](`r knitr::fig_chunk("simp-points", "png")`)
]


---
# Simple slopes

The interaction means that the *slope* of the effect of `stateanx` differs at different values of `epiNeur`.

We can use the `sim_slopes()` function from `interactions` to statistically explore how `stateanx` varies as a function of `epiNeur`.

```{r eval = FALSE}
sim_slopes(int_model,
           pred = stateanx,
           modx = epiNeur,
           johnson_neyman = FALSE)
```

---

```{r echo = FALSE}
sim_slopes(int_model,
           pred = stateanx,
           modx = epiNeur,
           johnson_neyman = FALSE)
```

The slope of `stateanx` *increases* as `epiNeur` increases.

---
# Johnson-Neyman plots

```{r fig.height=4, fig.width = 6}
johnson_neyman(int_model, pred = stateanx, modx = epiNeur)
```

---
# Moderation with the `psych` package

The function `mediate()` from the `psych` package can also be used to run moderation analyes.

.center[
```{r}
psych_mod <- mediate(bdi ~ stateanx * epiNeur,
                     data = epi.bfi)
```
]

---
# Moderation with the `psych` package

```{r}
psych_mod
```

---
# Moderation with the `medmod` package

We could also use the `mod()` function from `medmod`, which has rather nicer output than the `mediate()` function.

```{r}
mod(epi.bfi, dep = "bdi", mod = "epiNeur", pred = "stateanx")
```

---
class: inverse, center, middle
# First exercise

---
class: inverse, center, middle
# Mediation

---
# Mediation

Mediation refers to a situation in which the effect of a predictor is transmitted *through* another variable.

.center[
```{r echo = FALSE}
pmacroModel(4,
            radx = .1)
```
]

---
# Mediation path diagram

.pull-left[
```{r echo = FALSE}
statisticalDiagram(4)
```
]
.pull-right[
$a$ - the effect of the IV on the mediator

$b$ - the effect of the mediator on the DV

$c'$ - the *direct* effect of the IV on the DV 

What's missing is $c$ - the *total* effect of the IV on the DV - and $ab$ - the *indirect* effect of the IV on the DV
]


---
# Mediation

In this example, *room temperature* predicts the *amount that people drink*; specifically, we'd expect that higher temperatures would increase drinking.

.center[
```{r echo = FALSE}
openplotmat(asp = 1)
textrect(c(.1, .5), radx = .15, lab = "Room temp")
textrect(c(.9, .5), radx = .15, lab = "Amount drunk")
straightarrow(c(.25, .5), c(.75, .5))
```
]

---
# Mediation

Nevertheless, it's possible that higher temperatures increase drinking *indirectly*: higher temperatures make people feel more *thirsty*, which in turn makes them *drink more*.

.center[
```{r echo = FALSE}
pmacroModel(4,
            labels = list(X = "Room temp",
                          Mi = "Thirst",
                          Y = "Amount drunk"),
            radx = .1)
```
]


---
# Mediation as regression

A mediation analysis is all about splitting up the effect of the predictor into *direct* and *indirect* components. Baron & Kenny (1986) outline 3 steps to do this with regression.

###Step 1:

Does the IV predict the DV? (contrary to popular belief, doesn't have to be significant)

--
###Step 2:

Does the mediator predict the DV? (this needs to be significant)

--
###Step 3:

Does having the mediator in the model reduce the effect of the IV on the DV?


---
# The `estress` data

```{r}
head(estress)
```

---
# Mediation as regression (Step 1)

```{r highlight.output = 12}
summary(lm(withdraw ~ estress, data = estress))
```

---
# Mediation as regression (Step 2)

```{r highlight.output = 12}
summary(lm(withdraw ~ affect, data = estress))
```

---
# Mediation as regression (Step 3)

```{r highlight.output = 12}
summary(lm(withdraw ~ estress + affect, data = estress))
```

---
# Is there mediation?

**Step 1:**
`estress` wasn't a significant predictor of `withdraw`, but was a significant predictor of `affect`.

**Step 2:**
`affect` was a significant predictor of `withdraw`.

**Step 3:**
Modelling `withdraw` as a function of `estress` and `affect` changed the coefficient of `estress`.

Thus, we have evidence of mediation. How much mediation? For that we need the *indirect* effect.

The *indirect* effect of `estress` is the coefficient of `estress` from Step 1 MINUS the coefficient of `estress` from Step 3.

---
# Mediation model 

We can use the `mediate()` function from the `psych` package to add a mediating variable. **Importantly,** we place `()` around the mediator.

.pull-left[
```{r}
medi_model <- mediate(withdraw ~ estress + (affect), data = estress)
```
]
.pull-right[
The difference between $c'$ and $c$ is the *indirect* effect of `estress`.

When `estress` increases by 1, `affect` increases by .17; and when `affect` increases by 1, `withdraw` increases by .77.

So the *indirect effect* of `estress` is around **.14**.
]
---

```{r echo = FALSE}
summary(medi_model)
```

---
# Mediation with `med()` from `medmod`

```{r}
med_model <- med(data = estress, dep = "withdraw",
                 pred = "estress", med = "affect",
                 paths = TRUE, estPlot = TRUE)
med_model$med
```

---
# Mediation with `med()` from `medmod`

```{r}
med_model$paths
```

PS this output looks better direct from R...!

---
# Mediation with `med()` from `medmod`

.pull-left[
```{r message = FALSE, fig.height = 5}
med_model$estPlot
```
]
.pull-right[
As long as the confidence intervals don't overlap 0 for the indirect effect, we have a significant mediation.
]

---
class: inverse, center, middle
# Second exercise


---
# Further reading

Baron, R. M., & Kenny, D. A. (1986). The moderator–mediator variable distinction in social psychological research: Conceptual, strategic, and statistical considerations. *Journal of Personality and Social Psychology, 5, 1173-1182.*

Shrout, P. E., & Bolger, N. (2002). Mediation in experimental and nonexperimental studies: new procedures and recommendations. *Psychological Methods, 7, 422-445.*

Hayes AF. Introduction to mediation, moderation, and conditional process analysis: a regression-based approach. New York: Guilford Press; 2013.